Theory about Normal Basis on $F(2^k)$

### 2016-12-09 dissertation/abstract.tex
dissertation v2:add more to appendix update 3

\begin{center}{\bf ABSTRACT}\end{center}

Verification of sequential circuits remains to be a topic of all time. With increasing size of new
integrated circuits, sequential circuit designers may face much more complicated problems on 
logic errors, unexpected delays and failures on critical path. 
This dissertation addresses the problem of sequential circuit verification at the word-level
and is based on concepts derived from algebraic geometry.

Analyzing the sequential circuits at word level is an efficient way
of {\it abstraction}, which may lower the complexity of verification by efficient representation of the
state-space. We manage to model the verification properties and the gate-level sequential circuit 
implementations over Galois fields of the type ${\mathbb{F}}_{2^k}$ by means of polynomial ideals and 
their canonical representations --- Gr\"obner bases --- at the level of $k$-bit words. Subsequently, 
techniques from algebraic geometry can be used to reason about the state-space on high-level.
In algebraic geometry, ``bad" states are modeled as varieties while whole system is described using 
polynomial ideals. Without actually solving the system, the states in state space can be investigated 
by manipulating these ideals.

We propose to apply these techniques to traverse a finite state machine (FSM) for reachability 
analysis at the word-level, and also to implicitly unroll sequential arithmetic circuits to 
verify their function. Moreover, as unsatisfiable (UNSAT) cores play an important role in modern 
abstraction-refinement techniques for verification, we propose to investigate a word-level 
analogue of the UNSAT core problem using the  Gr\"obner basis algorithm. The dissertation will 
not only derive new algorithms and techniques, but will also consider efficient CAD 
implementations to target sequential equivalence and model checking problems. 
% \end{abstract}

>>>>

\begin{center}{\bf ABSTRACT}\end{center}
Formal verification of hardware designs has become an essential
component of the overall system design flow. The designs are generally
modeled as finite state machines, on which property and equivalence
checking problems are solved for verification. Reachability analysis
forms the core of these techniques. However, increasing
size and complexity of the circuits causes the state explosion
problem. Abstraction is key to tackle the scalability challenge.

This dissertation presents new techniques for word-level abstraction
with applications in sequential design verification. By bundling
together $k$ bit-level state-variables into one word-level constraint 
expression, the state-space is construed as solutions (variety) to
a set of polynomial constraints (ideal), modeled over the finite
(Galois) field of $2^k$ elements. Subsequently, techniques from
algebraic geometry -- notable, Gr\"obner basis theory and technology
-- are researched to perform reachability analysis and
verification of sequential circuits. This approach adds a ``word-level
dimension'' to state-space abstraction and verification to make the
process more efficient.

While algebraic geometry provides powerful abstraction and reasoning
capabilities, the algorithms exhibit high computational complexity. In
the dissertation, we show that by analyzing the constraints, it is
possible to obtain more insights about the polynomial ideals, which can be
exploited to overcome the complexity. Using our algorithm design and
implementations, we demonstrate how to perform reachability
analysis of finite-state machines purely at the word-level. Using this
concept, we perform scalable verification of sequential arithmetic
circuits. As contemporary approaches make use of resolution proofs and
unsatisfiable cores for state-space abstraction, we introduce the
algebraic geometry analog of unsatisfiable cores, and present
algorithms to extract and refine unsatisfiable cores of polynomial
ideals. Experiments are performed to demonstrate the efficacy of our
approaches.

### 2016-12-09 dissertation/appendix.tex
dissertation v2:add more to appendix update 3

\section{Optimal Normal Basis}
\label{append:ONB}
\subsection{Construction of Optimal Normal Basis}
\subsection{Optimal Normal Basis Multiplier Design}
% Customized Galois Field IEEE1363-2000
% Multi table <=> \lambda matrix
In Section \ref{sec:nbdesign}, we introduce the concept of multiplication table and explained how its 
entries are transformed from the $0$-th $\lambda$-Matrix. Actually we can mathematically prove the relationship 
between them:
\begin{Theorem}
Multiplication table $T$ is actually a conjugate of the $\lambda$-Matrix $M$. 
We can use an equation $M_{i,j}^{(0)} = T_{j-i,-i}$ to demonstrate the relation between their entries.
\end{Theorem}
\begin{Proof}
Recall the definition of $\lambda$-Matrix over field $\F_{2^k}$:
$$C = A\times B = \left(\sum_i a_i\beta^{2^i}\right)\left(\sum_jb_j\beta^{2^j}\right) = \sum_i\sum_ja_ib_j\beta^{2^i}\beta^{2^j}$$
Then there always exist $\lambda_{ij}^{(l)}$, such that 
$$\beta^{2^i}\beta^{2^j} = \sum_l\lambda_{ij}^{(l)}\beta^{2^l}$$
We call it the cross-product term. The $l$-th bit of the product $C$ is
$$c_l = \sum_i\sum_ja_ib_j\lambda_{ij}^{(l)}$$
For the sake of simplification, we let $i=0$, thus 
$$\beta\cdot\beta^{2^j} = \sum_l\lambda_{0j}^{(l)}\beta^{2^l}$$
Since the multiplication table is defined as
\begin{equation}
\beta
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^{2^2} \\ \vdots \\ \beta^{2^{n-1}}
\end{bmatrix}
= {\bf T}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^{2^2} \\ \vdots \\ \beta^{2^{n-1}}
\end{bmatrix}
\end{equation}
The $j$-th row of $T$ can be written as
\begin{align}
\beta\cdot\beta^{2^j} =& \sum_l\lambda_{0j}^{(l)}\beta^{2^l} & \text{(multi-table definition)} \nonumber\\
					=& \sum_l\lambda_{jl}^{(0)}\beta^{2^l} & \text{(cross-product term)} \label{eqn:mteq}
\end{align}
Notice that $\lambda$ corresponds to $\lambda$-Matrix entries. In Equation \ref{eqn:mteq} we assign $0$ to the 
row index $i$, the proof can be extended to all row indices $i<n$ but fix $l$ to 0 because of the 
conjugation generated by right-down cyclic-shift of $\lambda$-Matrix. Therefore we have 
$$M_{i,j}^{(0)} = \lambda_{ij}^{(0)} = \lambda_{j,0}^{(i)} = T_{j-i,0-i}$$
\end{Proof}

>>>>

\section{Optimal Normal Basis}
\label{append:ONB}
The number of non-zero entries in $\lambda$-Matrix or multiplication table is known as \emph{Complexity $(C_N)$}.
To define optimal normal basis, it is necessary to find the lower bound of $C_N$.
\begin{Theorem}
If N is a normal basis over $\F_{p^n}$ with $\lambda$-Matrix $M^{(k)}$, then non-zero entries in matrix $C_N\geq 2n-1$.
\end{Theorem}
\begin{Proof}
Let basis $N = \{\beta, \beta^p, \beta^{p^2},\dots, \beta^{p^{n-1}}\}$, let us denote $\beta^{p^i}$ by $\beta_i$ for the sake of simplification.
Then $\sum_{i=0}^{n-1} \beta_i = trace\ \beta$. 
Denote $trace\ \beta$ by $b$, consider a $n\times n$ matrix $M^{(0)}$. Then
$$b\beta_0 = \sum_{i=0}^{n-1} \beta \beta_i$$

Therefore, the sum of all rows in $M^{(0)}$ is an n-tuple with $b$ as the first element and zeros elsewhere.
So the first column always includes at least one nonzero element. 
Since the sum of enries in each of other columns should equal to zero (modulo 2), they 
need to include an even number of ``1"s.
in order to maintain the linear independence among each row, they cannot be purely ``0"s. 
Therefore, there are always at least two non-zero elements in each column.

As a result, $C_N \geq 2(n-1)+1  = 2n-1$.
\end{Proof}

If there exists a set of normal basis satisfying $C_N = 2n - 1$, this normal basis is named as \emph{Optimal Normal Basis}.

\begin{Example}
In $\F_{2^4}$ constructed with $p(\alpha) = \alpha^4 + \alpha + 1$, two sets of normal bases can be found:
$\beta = \alpha^3, N = \{ \alpha^3, \alpha^6, \alpha^{12}, \alpha^9\}$ and 
$\beta = \alpha^7, N = \{\alpha^7, \alpha^{14}, \alpha^{13}, \alpha^{11}\}$. Multiplication table for $\beta = \alpha^3$:
\begin{equation}
T_1 = \left(
\begin{array}{lccr}
0 & 1 & 0 & 0\\
0 & 0 & 0 & 1\\
1 & 1 & 1 & 1\\
0 & 0 & 1 & 0
\end{array} \right)
\end{equation}
Complexity $C_N$ = 7, this is optimal. For $\beta = \alpha^7$:
\begin{equation}
T_2 = \left(
\begin{array}{lccr}
0 & 1 & 0 & 0\\
1 & 1 & 0 & 1\\
1 & 0 & 1 & 0\\
1 & 0 & 1 & 1
\end{array} \right)
\end{equation}
Complexity $C_N$ = 9, this is NOT optimal.
\end{Example}

There should be an efficient method to construct optimal normal basis. The following  subsection refers to \cite{rosingbook}.
\subsection{Construction of Optimal Normal Basis}
\subsection{Optimal Normal Basis Multiplier Design}
% Customized Galois Field IEEE1363-2000
% Multi table <=> \lambda matrix
In Section \ref{sec:nbdesign}, we introduce the concept of multiplication table and explained how its 
entries are transformed from the $0$-th $\lambda$-Matrix. Actually we can mathematically prove the relationship 
between them:
\begin{Theorem}
Multiplication table $T$ is actually a conjugate of the $\lambda$-Matrix $M$. 
We can use an equation $M_{i,j}^{(0)} = T_{j-i,-i}$ to demonstrate the relation between their entries.
\end{Theorem}
\begin{Proof}
Recall the definition of $\lambda$-Matrix over field $\F_{2^k}$:
$$C = A\times B = \left(\sum_i a_i\beta^{2^i}\right)\left(\sum_jb_j\beta^{2^j}\right) = \sum_i\sum_ja_ib_j\beta^{2^i}\beta^{2^j}$$
Then there always exist $\lambda_{ij}^{(l)}$, such that 
$$\beta^{2^i}\beta^{2^j} = \sum_l\lambda_{ij}^{(l)}\beta^{2^l}$$
We call it the cross-product term. The $l$-th bit of the product $C$ is
$$c_l = \sum_i\sum_ja_ib_j\lambda_{ij}^{(l)}$$
For the sake of simplification, we let $i=0$, thus 
$$\beta\cdot\beta^{2^j} = \sum_l\lambda_{0j}^{(l)}\beta^{2^l}$$
Since the multiplication table is defined as
\begin{equation}
\beta
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^{2^2} \\ \vdots \\ \beta^{2^{n-1}}
\end{bmatrix}
= {\bf T}
\begin{bmatrix}
\beta \\ \beta^2 \\ \beta^{2^2} \\ \vdots \\ \beta^{2^{n-1}}
\end{bmatrix}
\end{equation}
The $j$-th row of $T$ can be written as
\begin{align}
\beta\cdot\beta^{2^j} =& \sum_l\lambda_{0j}^{(l)}\beta^{2^l} & \text{(multi-table definition)} \nonumber\\
					=& \sum_l\lambda_{jl}^{(0)}\beta^{2^l} & \text{(cross-product term)} \label{eqn:mteq}
\end{align}
Notice that $\lambda$ corresponds to $\lambda$-Matrix entries. In Equation \ref{eqn:mteq} we assign $0$ to the 
row index $i$, the proof can be extended to all row indices $i<n$ but fix $l$ to 0 because of the 
conjugation generated by right-down cyclic-shift of $\lambda$-Matrix. Therefore we have 
$$M_{i,j}^{(0)} = \lambda_{ij}^{(0)} = \lambda_{j,0}^{(i)} = T_{j-i,0-i}$$
\end{Proof}
